---
title: "Differential gene expression analysis"
output: html_notebook
---

Differential gene expression analysis will be carried out to get up and down regulated genes in the day and night periods.

# Importing data

Importing the matrix (same used in SparXCC, filtered based on CV and expression levels) and sample annotation.
Filtering steps will not be necessary, since it was previously carried out for SparXCC.

```{r}
cts <- as.matrix(read.csv("/home/rsantos/Repositories/maize_microbiome_transcriptomics/poster_presentation_notebooks/plant_center_retreat_2024/expression_counts_day_night_filtered_cv_filtered.tsv", sep="\t",row.names="Name"))
coldata <- read.csv("/home/rsantos/Repositories/maize_microbiome_transcriptomics/poster_presentation_notebooks/pag_2025/sample_annotation.txt", row.names=1, sep="\t")
```

Some checking steps and fixing column orders, as described in DESeq2 quick start:

```{r}
dim(cts)
dim(coldata)
head(cts)
head(coldata)
all(rownames(coldata) %in% colnames(cts)) # Check if all samples are present in the count matrix
all(rownames(coldata) == colnames(cts)) # Check if the order of samples is the same in the count matrix and the coldata
cts <- cts[, rownames(coldata)]
all(rownames(coldata) == colnames(cts))
```

Checking the column data. Object for DESeq2 will use DayPeriod as the design formula.

```{r}
head(coldata)
```

```{r}
library("DESeq2")
dds <- DESeqDataSetFromMatrix(countData = cts,
                              colData = coldata,
                              design = ~ DayPeriod)
dds
colData(dds)
```


# Differential gene expression analysis


```{r}
dds <- DESeq(dds)
res_day_periods_dea <- results(dds, contrast=c("DayPeriod","day","night"))
res_day_periods_dea
```


```{r}
resultsNames(dds)
resLFC <- lfcShrink(dds, coef="DayPeriod_night_vs_day", type="apeglm")
resLFC
```


```{r}
resOrdered <- res_day_periods_dea[order(res_day_periods_dea$pvalue),]
resOrdered
```

```{r}
summary(resOrdered)
sum(resOrdered$padj < 0.01, na.rm=TRUE)
```



```{r}
plotMA(resLFC, ylim=c(-4,4))
```

```{r}
plotCounts(dds, gene="Zm00001eb032990_T001", intgroup="DayPeriod")
```

## Multi-factor design formula

Using multi-factor design formula to account for the effect of the genotype:


```{r}
ddsMF <- dds # Making a copy of the DESeq2 object
colData(ddsMF)
colData(ddsMF)$Genotype <- as.factor(colData(ddsMF)$Genotype)
levels(ddsMF$Genotype) <- gsub("-", "_", levels(ddsMF$Genotype)) # Replacing hyphens by underscores in all their occurrences
levels(ddsMF$Genotype)
```

```{r}
design(ddsMF) <- formula(~ Genotype + DayPeriod)
ddsMF <- DESeq(ddsMF)
resMF <- results(ddsMF)
summary(resMF)
```


```{r}
head(resMF)
resMFOrdered <- resMF[order(resMF$pvalue),]
head(resMFOrdered)
plotCounts(ddsMF, gene="Zm00001eb256520_T003", intgroup="DayPeriod")
```