---
title: "Differential gene expression analysis"
output: html_notebook
---

Differential gene expression analysis will be carried out to get up and down regulated genes in the day and night periods.

# Importing data

Importing the matrix (same used in SparXCC, filtered based on CV and expression levels) and sample annotation.
Filtering steps will not be necessary, since it was previously carried out for SparXCC.

```{r}
cts <- as.matrix(read.csv("/home/rsantos/Repositories/maize_microbiome_transcriptomics/poster_presentation_notebooks/plant_center_retreat_2024/expression_counts_day_night.tsv", sep="\t",row.names="Name"))
coldata <- read.csv("/home/rsantos/Repositories/maize_microbiome_transcriptomics/poster_presentation_notebooks/pag_2025/sample_annotation.txt", row.names=1, sep="\t")
```

Some checking steps and fixing column orders, as described in DESeq2 quick start:

```{r}
dim(cts)
dim(coldata)
head(cts)
head(coldata)
all(rownames(coldata) %in% colnames(cts)) # Check if all samples are present in the count matrix
all(rownames(coldata) == colnames(cts)) # Check if the order of samples is the same in the count matrix and the coldata
cts <- cts[, rownames(coldata)]
all(rownames(coldata) == colnames(cts))
```

Checking the column data. Object for DESeq2 will use DayPeriod as the design formula.

```{r}
head(coldata)
```

# Running differential expression analysis with multi-factor design formula

Using multi-factor design formula to account for the effect of the genotypes:

```{r}
library("DESeq2")
ddsMF <- DESeqDataSetFromMatrix(countData = cts,
                              colData = coldata,
                              design = ~ DayPeriod)
ddsMF
colData(ddsMF)
colData(ddsMF)$Genotype <- as.factor(colData(ddsMF)$Genotype)
levels(ddsMF$Genotype) <- gsub("-", "_", levels(ddsMF$Genotype)) # Replacing hyphens by underscores in all their occurrences
design(ddsMF) <- formula(~ Genotype + DayPeriod) # Replacing the design formula to consider genotypes as a soruce of variation
```

Running the differential expression analysis:

```{r}
ddsMF <- DESeq(ddsMF)
resMF <- results(ddsMF)
summary(resMF)
```

```{r}
resultsNames(ddsMF)
resMF_LFC <- lfcShrink(ddsMF, coef="DayPeriod_night_vs_day", type="apeglm")
resMF_LFC
```


```{r}
head(resMF)
resMFOrdered <- resMF[order(resMF$pvalue),]
head(resMFOrdered)
resMFOrdered_filtered <- resMFOrdered[resMFOrdered$padj < 0.05,]
write.csv(as.data.frame(resMFOrdered), 
          file="day_night_results.csv")
write.csv(as.data.frame(resMFOrdered_filtered), 
          file="day_night_results_filtered_padj_0_05.csv")
```

```{r}
plotCounts(ddsMF, gene="Zm00001eb049640_T001", intgroup="DayPeriod")
```

Mapping between clock or clock-related genes in [Lai et al 2020](https://link.springer.com/article/10.1186/s12864-020-06824-3) and the maize v4 (used in transcript quantification) was done in Phytozome. Identifiers identified in maize v4 were mapped to IDs in maize genome v5 using a mapping file provided by MaizeGDB [here](https://download.maizegdb.org/Pan-genes/B73_gene_xref/).

```{r}
plotCounts(ddsMF, gene="Zm00001eb397240_T001", intgroup="DayPeriod")
plotCounts(ddsMF, gene="Zm00001eb122110_T003", intgroup="DayPeriod") # Not differentially expressed
plotCounts(ddsMF, gene="Zm00001eb228670_T002", intgroup="DayPeriod")
plotCounts(ddsMF, gene="Zm00001eb273260_T002", intgroup="DayPeriod") # Not differentially expressed
plotCounts(ddsMF, gene="Zm00001eb274210_T002", intgroup="DayPeriod") # Not differentially expressed
plotCounts(ddsMF, gene="Zm00001eb379300_T001", intgroup="DayPeriod") # Not differentially expressed
plotCounts(ddsMF, gene="Zm00001eb389960_T001", intgroup="DayPeriod") # Not differentially expressed
```

