---
title: "Functional analyses of bipartite networks (PAG 2025)"
output: html_notebook
---

Installing the `clusterProfiler` package from Bioconductor.

```{r}
# Installation step
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install(version = "3.20")
BiocManager::install("clusterProfiler")
library(clusterProfiler)
```

Enrichment analysis of the bipartite network for KEGG pathways.
It looks like there is maize among the available organisms in KEGG; however, the identifiers are not the same as the ones I have been working with.

```{r}
search_kegg_organism('zma', by='kegg_code')
zma <- search_kegg_organism('Zea mays', by='scientific_name')
dim(zma)
zma
```

It is possible to use KOs as identifiers. Since KOs are available for maize genes on Phytozome, I will use them to perform the enrichment analysis.

```{r}
# Importing a list of identifiers from a file and converting to a vector
day_all_identifiers <- read.table("/home/santosrac/Projects/UGA_RACS/IntegrationMetataxExpression/SparXCC_OTUs_pag2025/modules/day_mod_all_kos.txt", header = FALSE, stringsAsFactors = FALSE)
day_all_kos <- as.vector(day_all_identifiers$V1)  # Assuming the identifiers are in the first column
length(day_all_kos)
day_1_identifiers <- read.table("/home/santosrac/Projects/UGA_RACS/IntegrationMetataxExpression/SparXCC_OTUs_pag2025/modules/day_mod_1_kos.txt", header = FALSE, stringsAsFactors = FALSE)
day_1_kos <- as.vector(day_1_identifiers$V1)  # Assuming the identifiers are in the first column
length(day_1_kos)
day_2_identifiers <- read.table("/home/santosrac/Projects/UGA_RACS/IntegrationMetataxExpression/SparXCC_OTUs_pag2025/modules/day_mod_2_kos.txt", header = FALSE, stringsAsFactors = FALSE)
day_2_kos <- as.vector(day_2_identifiers$V1)  # Assuming the identifiers are in the first column
length(day_2_kos)
day_3_identifiers <- read.table("/home/santosrac/Projects/UGA_RACS/IntegrationMetataxExpression/SparXCC_OTUs_pag2025/modules/day_mod_3_kos.txt", header = FALSE, stringsAsFactors = FALSE)
day_3_kos <- as.vector(day_3_identifiers$V1)  # Assuming the identifiers are in the first column
length(day_3_kos)
day_4_identifiers <- read.table("/home/santosrac/Projects/UGA_RACS/IntegrationMetataxExpression/SparXCC_OTUs_pag2025/modules/day_mod_4_kos.txt", header = FALSE, stringsAsFactors = FALSE)
day_4_kos <- as.vector(day_4_identifiers$V1)  # Assuming the identifiers are in the first column
length(day_4_kos)
day_5_identifiers <- read.table("/home/santosrac/Projects/UGA_RACS/IntegrationMetataxExpression/SparXCC_OTUs_pag2025/modules/day_mod_5_kos.txt", header = FALSE, stringsAsFactors = FALSE)
day_5_kos <- as.vector(day_5_identifiers$V1)  # Assuming the identifiers are in the first column
length(day_5_kos)
day_6_identifiers <- read.table("/home/santosrac/Projects/UGA_RACS/IntegrationMetataxExpression/SparXCC_OTUs_pag2025/modules/day_mod_6_kos.txt", header = FALSE, stringsAsFactors = FALSE)
day_6_kos <- as.vector(day_6_identifiers$V1)  # Assuming the identifiers are in the first column
length(day_6_kos)
```

```{r}
enrichKEGG(
  day_all_kos,
  organism = "ko",
  keyType = "kegg",
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH",
  minGSSize = 10,
  maxGSSize = 500,
  qvalueCutoff = 0.2,
  use_internal_data = FALSE
)

enrichKEGG(
  day_1_kos,
  organism = "ko",
  keyType = "kegg",
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH",
  minGSSize = 10,
  maxGSSize = 500,
  qvalueCutoff = 0.2,
  use_internal_data = FALSE
)

enrichKEGG(
  day_2_kos,
  organism = "ko",
  keyType = "kegg",
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH",
  minGSSize = 10,
  maxGSSize = 500,
  qvalueCutoff = 0.2,
  use_internal_data = FALSE
)

enrichKEGG(
  day_3_kos,
  organism = "ko",
  keyType = "kegg",
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH",
  minGSSize = 10,
  maxGSSize = 500,
  qvalueCutoff = 0.2,
  use_internal_data = FALSE
)

enrichKEGG(
  day_4_kos,
  organism = "ko",
  keyType = "kegg",
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH",
  minGSSize = 10,
  maxGSSize = 500,
  qvalueCutoff = 0.2,
  use_internal_data = FALSE
)

enrichKEGG(
  day_5_kos,
  organism = "ko",
  keyType = "kegg",
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH",
  minGSSize = 10,
  maxGSSize = 500,
  qvalueCutoff = 0.2,
  use_internal_data = FALSE
)

enrichKEGG(
  day_6_kos,
  organism = "ko",
  keyType = "kegg",
  pvalueCutoff = 0.05,
  pAdjustMethod = "BH",
  minGSSize = 10,
  maxGSSize = 500,
  qvalueCutoff = 0.2,
  use_internal_data = FALSE
)
```