---
title: "SparXCC PAG 2025"
output: html_notebook
---

Importing SparXCC library:

```{r}
library(devtools)
#install_github("IbTJensen/CompoCor")
library(CompoCor)
```

Importing the tables:

```{r}
# Day samples
#wallace_otu_counts_filtered_day <- read.csv("C:/Users/renat/Downloads/tmp/pag25/filtered_otu_table_day_filtered_rel_abund_cv_filtered.tsv", sep="\t", header=TRUE, row.names=1)
wallace_otu_counts_filtered_day <- read.csv("/home/santosrac/Repositories/maize_microbiome_transcriptomics/poster_presentation_notebooks/pag_2025/filtered_otu_table_day_filtered_rel_abund_cv_filtered.tsv", sep="\t", header=TRUE, row.names=1)
dim(wallace_otu_counts_filtered_day)
head(wallace_otu_counts_filtered_day)
colnames(wallace_otu_counts_filtered_day) <- gsub("^X", "", colnames(wallace_otu_counts_filtered_day))

# Night samples
#wallace_otu_counts_filtered_night <- read.csv("C:/Users/renat/Downloads/tmp/pag25/filtered_otu_table_night_filtered_rel_abund_cv_filtered.tsv", sep="\t", header=TRUE, row.names=1)
wallace_otu_counts_filtered_night <- read.csv("/home/santosrac/Repositories/maize_microbiome_transcriptomics/poster_presentation_notebooks/pag_2025/filtered_otu_table_night_filtered_rel_abund_cv_filtered.tsv", sep="\t", header=TRUE, row.names=1)
dim(wallace_otu_counts_filtered_night)
head(wallace_otu_counts_filtered_night)
colnames(wallace_otu_counts_filtered_night) <- gsub("^X", "", colnames(wallace_otu_counts_filtered_night))

#kremling_expression_counts_filtered <- read.csv("C:/Users/renat/Downloads/tmp/pag25/expression_counts_day_night_filtered_cv_filtered.tsv", sep="\t", header=TRUE, row.names=1)
kremling_expression_counts_filtered <- read.csv("/home/santosrac/Projects/UGA_RACS/IntegrationMetataxExpression/SparXCC_summed_OTUs_plant_center_retreat/expression_counts_day_night_filtered_cv_filtered.tsv", sep="\t", header=TRUE, row.names=1)
head(kremling_expression_counts_filtered)

kremling_expression_counts_filtered_day <- kremling_expression_counts_filtered[, grep("day", names(kremling_expression_counts_filtered), value = TRUE)]
kremling_expression_counts_filtered_night <- kremling_expression_counts_filtered[, grep("night", names(kremling_expression_counts_filtered), value = TRUE)]

head(kremling_expression_counts_filtered_day)
dim(kremling_expression_counts_filtered_day)
colnames(kremling_expression_counts_filtered_day) <- gsub("day_", "", colnames(kremling_expression_counts_filtered_day))
head(kremling_expression_counts_filtered_day)

head(kremling_expression_counts_filtered_night)
dim(kremling_expression_counts_filtered_night)
colnames(kremling_expression_counts_filtered_night) <- gsub("night_", "", colnames(kremling_expression_counts_filtered_night))
head(kremling_expression_counts_filtered_night)

wallace_otu_counts_filtered_day_sorted <- wallace_otu_counts_filtered_day[ , colnames(kremling_expression_counts_filtered_day)]
wallace_otu_counts_filtered_night_sorted <- wallace_otu_counts_filtered_night[ , colnames(kremling_expression_counts_filtered_night)]
```

Transposing dataframes:

```{r}
wallace_otu_counts_filtered_day_transposed <- t(wallace_otu_counts_filtered_day_sorted)
wallace_otu_counts_filtered_night_transposed <- t(wallace_otu_counts_filtered_night_sorted)
kremling_expression_counts_filtered_day_transposed <- t(kremling_expression_counts_filtered_day)
kremling_expression_counts_filtered_night_transposed <- t(kremling_expression_counts_filtered_night)
```

Making sure row names (samples) are identical between OTU and Expression data:

```{r}
identical(row.names(wallace_otu_counts_filtered_day_transposed), row.names(kremling_expression_counts_filtered_day_transposed))
identical(row.names(wallace_otu_counts_filtered_night_transposed), row.names(kremling_expression_counts_filtered_night_transposed))
```

```{r}
SparXCC_output_day <- SparXCC_base(wallace_otu_counts_filtered_day_transposed,
                        kremling_expression_counts_filtered_day_transposed,
                        pseudo_count = 1,
                        var_min = 1e-05,
                        Find_m = TRUE,
                        B_m = 100,
			cores = 1)

write.table(SparXCC_output_day, file = "SparXCC_output_day_pag25.txt", sep = "\t")

SparXCC_output_night <- SparXCC_base(wallace_otu_counts_filtered_night_transposed,
                        kremling_expression_counts_filtered_night_transposed,
                        pseudo_count = 1,
                        var_min = 1e-05,
                        Find_m = TRUE,
                        B_m = 100,
			cores = 1)

write.table(SparXCC_output_night, file = "SparXCC_output_night_pag25.txt", sep = "\t")
```

Since we noticed that the difference of numbers of cross-correlations between day and night samples is too high, we decided to run the SparXCC with the same plots for both day and night.

```{r}
# Find intersection of OTUs between day and night samples
common_plots <- intersect(colnames(wallace_otu_counts_filtered_day), colnames(wallace_otu_counts_filtered_night))
common_plots

# Select columns based on vector
wallace_otu_counts_filtered_day_common_plots <- wallace_otu_counts_filtered_day[ , common_plots]
wallace_otu_counts_filtered_night_common_plots <- wallace_otu_counts_filtered_night[ , common_plots]
kremling_expression_counts_filtered_day_common_plots <- kremling_expression_counts_filtered_day[ , common_plots]
kremling_expression_counts_filtered_night_common_plots <- kremling_expression_counts_filtered_night[ , common_plots]

head(wallace_otu_counts_filtered_day_common_plots)
head(kremling_expression_counts_filtered_day_common_plots)

# Check if columns are the same and in the same order
identical(colnames(wallace_otu_counts_filtered_day_common_plots), colnames(kremling_expression_counts_filtered_day_common_plots))
identical(colnames(wallace_otu_counts_filtered_night_common_plots), colnames(kremling_expression_counts_filtered_night_common_plots))

# Transpose dataframes
wallace_otu_counts_filtered_day_common_plots_transposed <- t(wallace_otu_counts_filtered_day_common_plots)
wallace_otu_counts_filtered_night_common_plots_transposed <- t(wallace_otu_counts_filtered_night_common_plots)
kremling_expression_counts_filtered_day_common_plots_transposed <- t(kremling_expression_counts_filtered_day_common_plots)
kremling_expression_counts_filtered_night_common_plots_transposed <- t(kremling_expression_counts_filtered_night_common_plots)
```

Running SparXCC with the same plots for both day and night samples:

```{r}
SparXCC_output_common_plots_day <- SparXCC_base(wallace_otu_counts_filtered_day_common_plots_transposed,
                        kremling_expression_counts_filtered_day_common_plots_transposed,
                        pseudo_count = 1,
                        var_min = 1e-05,
                        Find_m = TRUE,
                        B_m = 100,
			cores = 8)

write.table(SparXCC_output_common_plots_day, file = "SparXCC_output_day_common_samples_pag25.txt", sep = "\t")

SparXCC_output_common_plots_night <- SparXCC_base(wallace_otu_counts_filtered_night_common_plots_transposed,
                        kremling_expression_counts_filtered_night_common_plots_transposed,
                        pseudo_count = 1,
                        var_min = 1e-05,
                        Find_m = TRUE,
                        B_m = 100,
			cores = 8)

write.table(SparXCC_output_common_plots_night, file = "SparXCC_output_night_common_samples_pag25.txt", sep = "\t")
```